---
title: |
  High dose inactivated influenza vaccine inconsistently improves heterologous antibody responses in an elderly human cohort
author: " "
format:
  docx: 
    toc: false
    number-sections: false
    reference-doc: "template_segoe.docx"
bibliography:
  - "refs.bib"
  - "pkgs.bib"
  - "mybib.bib"
csl: "nlm_csl.csl"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	error = FALSE,
	knitr.graphics.error = FALSE
)
library(yaml, include.only = NULL)
library(knitr, include.only = NULL)
library(flextable)
library(softbib)

# Loading data to present in text
cate <- readr::read_rds(here::here("results", "files", "cate-formatted.Rds"))
```

**Authors**  

* W. Zane Billings$^{1,2,\land,*}$ (ORCID: 0000-0002-0184-6134);
* Yang Ge$^{3, \land, *}$ (0000-0001-5100-0703);
* Jessica H. Knight$^{1}$ (0000-0001-5764-7495);
* Hayley Hemme$^{1}$ (0009-0002-6609-1390);
* Savannah M. Hammerton$^{1, 2}$ (0000-0003-2231-3510);
* Amanda L. Skarlupka$^{4, \dagger}$ (0000-0002-3654-9076);
* Wangnan Cao$^{5}$ (0000-0002-6163-2760);
* Ye Shen$^{1}$;
* Justin Bahl$^{1, 2}$ (0000-0001-7572-4300);
* Paul G. Thomas$^{6}$;
* Ted M. Ross$^{7,8}$ (0000-0003-1947-7469);
* Andreas Handel$^{1, 2, \land}$ (0000-0002-4622-1146)

**Author affiliations**  

1. College of Public Health, University of Georgia, Athens, GA, USA.
2. Center for the Ecology of Infectious Diseases, The University of Georgia, Athens, GA, USA.
3. School of Health Professions, University of Southern Mississippi, Hattiesburg, MS, USA.
4. National Cancer Institute, Bethesda, MD, USA.
5. School of Public Health, Peking University, Beijing, China.
6. St. Jude Children's Research Hospital, Memphis, TN, USA.
7. Center for Vaccines and Immunology, University of Georgia, Athens, GA, USA.
8. Florida Research \& Innovation Center, Cleveland Clinic, Port St. Lucie, FL, USA.

$*$ These authors (WZB and YG) contributed equally to this work.

$\land$ Co-corresponding authors: WZB (wesley.billings@uga.edu), YG (Yang.Ge@usm.edu), AH (ahandel@uga.edu)

$\dagger$ Disclaimer: This article was prepared while Amanda L. Skarlupka was employed
at University of Georgia. The opinions expressed in this article are the
author's own and do not reflect the view of the National Institutes of Health,
the Department of Health and Human Services, or the United States government.

\pagebreak

# Abstract

**Background:** Older adults often mount a weak immune response to standard inactivated influenza vaccines. To induce a stronger response and better protection, a high-dose (HD) version of the
inactivated Fluzone vaccine is recommended for individuals >65 years of age.
While better immunogenicity and protection against the vaccine strain has been
shown, it is not known if the HD vaccine also induces a better antibody response
to heterologous strains.

**Methods:** We fit bayesian multilevel regression models to hemagglutination inhibition (HAI) antibody data from an influenza vaccine cohort spanning the 2013/14-2021/22 influenza seasons. We used this model to estimate the average causal effect (ACE) of obtaining the HD vaccine, relative to the SD vaccine.

**Results:** We show that while there is generally a benefit derived from the HD vaccine, the impact is small and inconsistent. For some strains, the HD vaccine might even result in less robust heterologous responses.

**Conclusions:** We suggest that further increases in dose might be worth investigating to help
induce a stronger broad response.

**Keywords:** influenza, high-dose influenza vaccine, heterologous responses

\pagebreak

# Introduction

Influenza vaccines are an important tool for reducing the burden of seasonal
influenza, but the average effectiveness is often less than 50\%
[@centersfordiseasecontrolandprevention2021seasons]. Standard dose (SD) split
inactivated vaccines, such as Sanofi Pasteur's Fluzone, comprise the majority of
vaccines licensed in the US [@CDCVaccineList], and are formulated to contain 15
micrograms of influenza hemagglutinin (HA) antigen protein for each strain of
influenza included in the vaccine [@FluzoneSDInsert]. Influenza vaccines often
induce a weak response in elderly individuals [@grohskopf2021prevention;
@robertson2016a]. In response, Sanofi Pasteur developed a high dose (HD)
formulation of the vaccine, Fluzone HD, which contains 60 micrograms of HA per
strain in one dose [@centersfordiseasecontrolandpreventioncdc2010licensure;
@falsey2009randomized].

In elderly individuals, HD vaccines induce a stronger homologous antibody
response compared to SD vaccines to the influenza strains contained in the
vaccine  [@couch2007safety;
@falsey2009randomized; @diazgranados2013highdose; @diazgranados2014efficacy;
@lee2021efficacy; @chang2019a; @chaves2023]. Additionally, HD vaccination is
associated with reduced disease severity and reduced risk of complications in
elderly individuals who contract influenza after vaccination [@chaves2023]. HD
vaccines also have the potential to elicit stronger immune responses in younger
individuals [@gouma2020], but younger individuals can develop protective immune
responses with SD vaccine, and even fractional doses of SD vaccine [@lunny2021;
@schnyder2020]. Therefore, the HD vaccine is currently only recommended for
elderly individuals.

While several studies have shown the ability of HD vaccines to induce stronger
antibody responses to the HA contained in the vaccine, whether HD vaccines
induce a stronger heterologous antibody responses (cross-reactive antibodies to
strains not included in the vaccine formulation) is uncertain.
[@couch2007safety; @hilleman1958antibody] Since our ability to forecast which
strains of influenza will circulate in the upcoming season is imperfect, it is
important for influenza vaccines to induce both homologous and heterologous
responses [@erbelding2018universal; @paules2018chasing].

It is not obvious if HD vaccines are expected to have a positive, negative or no
effect on heterologous responses. A higher dose might stimulate multiple diverse
lineages of memory B-cells, and could enhance both the strength and breadth of
protection. However, a higher amount of HA proteins to the homologous strains
could induce a stronger homologous response, which might out-compete and
immuno-dominate responses to any cross-reactive or novel antigens, resulting in
less robust heterologous responses. The potential competition among HD
vaccine-induced B-cells or T-cells might lead to more narrow protection from
vaccination [@angeletti2018understanding], making high-risk populations more
vulnerable to vaccine mismatches.

Improving our understanding of the heterologous immune response to influenza
vaccination remains a critical step in developing a broadly reactive influenza
vaccine. In our study, we compare the antibody response between HD and SD
vaccine recipients in a vaccine cohort study, using a panel of several
historical influenza A strains. We found that while the HD response was higher
for most strains, the impact was not consistent and the effect size was small,
suggesting that HD vaccines could be further improved.

# Methods

## Data source

We used data from an ongoing human vaccination cohort study, which has been
previously described in detail [@nunez2017impact; @carlock2019impact;
@abreu2020iga]. The study is a prospective open cohort which allows individuals
to enroll in multiple years (potentially non-consecutively), and has been
conducted across three study sites. Investigators annually recruited individuals
who had not yet received an influenza vaccine in the current season. At intake,
individuals provided demographic information, and investigators collected blood
samples before administering a vaccine. Individuals aged 65 and older were given
the choice between Fluzone SD and Fluzone HD, while individuals under 65 were
given the SD vaccine. Individuals were asked to return for post-vaccination
blood draws either 21 days (2013/14 through 2017/18 seasons) or 28 days (2018/19 season onward) after the initial
visit. Investigators ran hemagglutination inhibition (HAI) assays the vaccine
strains and several historical influenza strains for each serological sample.

For our analysis, we extracted previously deidentified records for individuals
aged 65 and older from 2013/2014 influenza season through the 2021/2022
influenza season.


## Data processing

The raw data for our study were reciprocal HAI titer values produced by the HAI
dilution assays described in the previous studies, along with limited
demographic information for each participant (study site, season, age, sex,
race, and a numeric identifier to track individuals across participation years).
Since the SD vaccine was quadrivalent (containing two influenza B lineages)
while the HD vaccine was trivalent (containing only a single influenza B
lineage) for multiple years of the study, we elected to focus only on the
heterologous response to influenza A virus strains, and excluded all data for
influenza B strains.

The data represent a multilevel structure, where each study site recruited
individuals, each individual could participate in multiple seasons, in each
season an individual had a pre-vaccination and a post-vaccination serological
sample, and investigators ran a panel of HAI assays to multiple strains on each
serological sample. The HAI assays had a lower limit of detection (LoD) of 10
and an upper LoD of 20480. Values below the lower LoD were coded as 5 in the raw
data. No values in our dataset were at the upper LoD. Following previous
studies, [@beyer2004seroprotection; @ranjeva2019agespecific] we conducted all
analyses on the log scale, using the transformation
$$y = \log_2\left(\frac{\text{raw reciprocal HAI titer}}{5}\right),$$
where $y$ is the titer variable we analyze. Using this transformation puts the
titers on a scale of $0$ (values at the lower LoD) $, 1, 2, \ldots, 12$ (values
at the upper LoD).

## Outcome definitions

We calculated four common outcomes used in influenza vaccine studies
[@beyer2004seroprotection; @falsey2009randomized]. The primary outcome we used
for our study was titer increase, defined as the log (base 2) fold change
between an individual's post-vaccination and pre-vaccination titer. That is,
$$
\text{titer increase} = \log_2 \left(\frac{y_\text{post}}{y_\text{pre}}\right) = \log_2\left(y_{\text{post}}\right) - \log_2\left(y_{\text{pre}}\right).
$$

We present results for titer increase in the main manuscript. As a sensitivity
analysis, we repeated our analysis using three additional outcomes:
post-vaccination titer, seroprotection, and seroconversion. Seroprotection and
seroconversion are binary outcomes, and this dichotomization leads to a loss of
statistical power. However, these are commonly reported measures and can be
useful for their clinical interpretation. We used standard definitions for
seroprotection and seroconversion, with seroprotection defined as a
post-vaccination titer greater than 1:40, that is,
$$
\mathrm{seroprotection} = I(\text{post-vaccination HAI
titer} \geq 40),
$$
where $I(\cdot)$ is the indicator function. The definition of seroconversion we
used was seroprotection in addition to a
titer increase of 2 or higher (i.e., a fold change of 4-fold or higher), that is,
$$
\mathrm{seroconversion} =
\text{seroprotection } \times \ I(\text{titer increase} \geq 2).
$$
Results for these three additional outcomes are included in the Supplement.

## Statistical analyses

For an initial descriptive analysis of the study population, we calculated
summary statistics for the covariates we included in our models and for each of
the model outcomes previously described. Summary statistics for the outcome
variables, including analyses stratified by vaccine strain and assay strain, are
shown in the Supplement. We conducted a crude analysis of the difference in fold
change by dose for each vaccine strain and assay strain, also shown in the
Supplement.

For our main results, we fit bayesian multilevel regression models for each of
the outcomes [@faraway2016extending; @mcelreath2020statistical]. Bayesian
multilevel regression can estimate the average effect of dose in the population,
while allowing the effect to vary across strata and flexibly controlling for
nuisance confounders. We allowed the effect of dose to vary by vaccine strain
and HAI assay strain, and effects for different groups were assumed to be
correlated with a heterogeneous unstructured covariance matrix. We used random
intercepts for subject and study site to control for baseline differences, and
used smoothing splines to control for the effects of birth year, age,
pre-vaccination titer, and season.

We used these models to estimate the effect of dose on each outcome of interest
by calculating the posterior Average Causal Effect (ACE). To do so, we compared
each individual's predicted outcome under the dose received (HD or SD) to their
predicted outcome for the counterfactual dose computed counterfactual HD and SD
predictions for each individual, thereby estimating the individual causal effect
(ICE) for each observation in the dataset. We estimated the ACE by summarizing
the overall posterior distribution of ICEs using the mean point estimate with a
95\% highest density credible interval (HDCI). We also computed conditional ACEs
(cACEs) by summarizing the posterior distribution of ICEs in specific strata:
within each season, within each vaccine strain, and within each assay strain and
vaccine strain. In order to aid interpretation, we then base-2 exponentiated the
ACEs and the limits of the HDCI. Each ACE represents the average difference in
the model outcome across the full study sample between the high dose and
standard dose vaccines, after controlling for confounders [@holland1986;
@naimi2023]. See the Supplement for additional details on cACE calculation.

Our study is a secondary data analysis, and the data were not collected with our
specific research questions in mind. Therefore, a strict hypothesis testing
framework using $p$-values is not appropriate. Any statistical tests we conduct
will have inflated false discovery rates and limited power. Therefore, we
implemented models in a bayesian framework, and we focus on estimating the
effect size and uncertainy as captured by credible intervals of the dose in our
dataset, rather than testing hypotheses.

Details on model specification, outcome calculation, and details of
implementation, including full data and code to reproduce results, are included
in the Supplement.

## Implementation

We conducted all of our analyses using `r version$version.string` [@base]. We
used the `tidyverse` suite of packages for data cleaning, manipulation, and
visualization [@tidyverse], along with the packages `tidybayes` [@tidybayes] and
`ggdist` [@ggdist1; @ggdist2]. We used the packages `here` [@here], `renv`
[@renv], and `box` [@box] for code management. We used `ggplot2` for generating
all figures [@ggplot2]. We used the packages `gtsummary` [@gtsummary] and
`flextable` [@flextable] for making all tables. We compiled our manuscript using
Quarto version 1.5.43 [@Allaire_Quarto_2024] with the R packages `knitr`
[@knitr1;
@knitr2; @knitr3], and `softbib` [@softbib]. We implemented our bayesian models
using `brms` [@brms1; @brms2; @brms3] and `cmdstanr` [@cmdstanr] with cmdstan
version 2.34.1 [@gabry2023] as the interface to the Stan probabilistic programming
language [@standevelopmentteam2022; @carpenter2017a]. More details on
implementation are included in the Supplement, along with instructions for
reproducing our results. The code and data are archived on Zenodo at this link:
[https://doi.org/10.5281/zenodo.12666977](https://doi.org/10.5281/zenodo.12666977).

# Results

## Data description

```{r}
#| label: load tot_n

tot_n <- readr::read_rds(here::here("results", "files", "n-indiv.Rds"))
```

Our data come from a prospective open cohort study and span the influenza
seasons from 2013/14 through 2021/22, and included `r tot_n` unique individuals
across all study sites and seasons, who provided 668 total observations.
Participants aged 65 and older were vaccinated with either Fluzone SD or HD at
one of three study sites (Stuart, FL and Pittsburgh, PA, from 2013--2016 and the
University of Georgia (UGA) in Athens, GA from 2016--2021). @tbl-demographics
shows the distribution of observations, stratified by vaccine dose. The three
study sites had similar distributions of demographic characteristics at both the
observation and unique participant levels (see Supplement). Since we only
include individuals 65 or older in our study, the age ranges and birth cohorts
were similar for both dose groups. Throughout the course of the study, the
recommended influenza vaccine composition was updated several times,
encompassing 5 different H1N1 strains and 8 different H3N2 strains.

```{r tbl-demographics}
#| label: tbl-demographics
#| tbl-cap: |
#|   Number of observations in our sample, stratified by the study site.
readr::read_rds(here::here("results", "tables", "demographics.Rds")) |>
	flextable::autofit()
```

## Strain-specific effects reveal differences in vaccine response patterns

Our models allowed the effect of dose to vary by the assay strain used for HAI
assays, so we first examined the strain-specific effects of dose.

For the H1N1 vaccine component, the titer increase was higher for HD in all of
the H1N1 strains except for CA/09 (@fig-all-strain-cates A), where the point
estimates for the heterologous responses were a mix of small positive and small
negative values. MI/15 had positive point estimates for nearly all of the assay
strains. Bris/18 showed a negative impact of dose for the single older strain
used for testing, and positive impact for the other strains. GD/19 and Vic/19
had positive point estimates with almost all of the density of the interval
estimate above 1 for the few strains they were tested against. The effect of the
HD vaccine on the Chi/83 strain was negative for all three of the vaccines which
it was tested against.

For H3N2, the effect of dose on heterologous responses showed patterns that
varied based on vaccine strain (@fig-all-strain-cates B). For TX/12 and
Switz/13, dose generally had a positive effect and this tended to increase as
the year of isolation for the assay strain became more similar to the homologous
strain. HK/14 showed an overall similar pattern, but with point estimates that
suggested a negative dose effect for older strains. Nearly every assay strain
for the Sing/16 and Tas/20 vaccines showed a negative effect of HD on
heterologous and homologous responses. For KS/17 and HK/19, dose seemed to have
an overall small positive impact.


```{r}
#| label: fig-all-strain-cates
#| fig-cap: |
#|   Strain-specific ratio of HD over SD antibody responses of heterologous
#|   strains following vaccination with the indicated vaccine strain. Values
#|   above one indicate a better response for the HD vaccine, values below
#|   one indicate a better response for the SD vaccine. A) H1N1 vaccine
#|   strains. B) H3N2 vaccine strains. For more recent vaccines, only a subset
#|   of data for heterologous responses was available.
knitr::include_graphics(
	here::here("Results", "Figures", "all-strains-cate.png")
)
```

## Most, but not all, HD vaccine formulations elicited a stronger overall response

```{r get overall effect}
# Pull the overall effect from the CACE data frame
overall_cate <-
	cate |>
	dplyr::filter(
		cate_label == "overall",
		model_label == "Titer increase",
		data_label == "heterologous"
	) |>
	dplyr::select(y, ymin, ymax) |>
	dplyr::summarize(
		est = paste0(
			sprintf("%.2f", y), "\\; 95\\% HDCI: (", sprintf("%.2f", ymin), ", ",
			sprintf("%.2f", ymax), ")"
		)
	) |>
	dplyr::pull(est)
```

After examining the strain-specific impact of the HD vaccine, we computed the
overall impact of the dose on each vaccine strain, by essentially computing the
difference in average response across all tested strains. We found a weakly
positive overall effect of the HD vaccine for all H1N1 strains (@fig-het-vex).
While the uncertainty in our parameter estimates is large, all of the HDCI
estimates were consistent with a small positive benefit from the HD vaccine. For
the H3N2 strains, the majority of vaccine strains showed a benefit for the HD
vaccine. However, the HK/14, Sing/16, and Tas/20 vaccines all had negative point
estimates.


```{r}
#| label: fig-het-vex
#| fig-cap: |
#|   Ratio of HD over SD average responses following vaccination with the
#|   indicated vaccine strain.
knitr::include_graphics(
	here::here("results", "figures", "heterologous-vaccine-cate.png")
)
```

## HD vaccines elicited stronger or equal responses in every influenza season.

For our final analysis, we pooled the H1N1 and H3N2 responses for a given
vaccine and computed an overall effect of the dose for a given vaccine in each
season (ignoring the influenza B components).

We found that an increase in dose had a positive but small impact for most
seasons, while for a few seasons, we did not observe an impact
(@fig-season-cates). On a vaccine-level basis, there was no indication of an
overall detrimental effect of dose.

```{r}
#| label: fig-season-cates
#| fig-cap: |
#|   Ratio of HD over SD average responses across all influenza A strains for
#|   a given vaccine and all heterologous responses.

knitr::include_graphics(
	here::here("Results", "Figures", "season-only-cate.png")
)
```


# Discussion

We analyzed homologous and heterologous antibody responses to influenza A HD and
SD vaccines across multiple years of influenza vaccinations.

While our estimates had high uncertainties and our secondary analyses should be
interpreted as exploratory, our results preliminarily suggest a small but
positive effect of the HD split-inactivated vaccine compared to the SD vaccine
for inducing not only homologous (similar to prior results; see Supplement) but
also better heterologous responses. However, this was not consistent and for
some vaccine strains, there was a trend towards a negative impact for the HD
vaccine. In general, both positive and negative effect sizes were small and most
of the time the credible intervals included both a no-effect as well as positive
and negative effect regions. Given that our analysis is a secondary data
analysis of noisy observational data, and the effect sizes (increases in titer)
for both SD and HD vaccines in our fairly heavily vaccinated population are not
that large, the amount of uncertainty we observed is expected.

Our results suggest that the HD vaccine does induce both better homologous and
heterologous responses for the majority of vaccine strains and thus should be a
continued recommendation for older individuals.

Our results also suggest that the overall impact of the HD vaccine is very
modest at inducing stronger antibody responses. Since we only examined the
humoral immune response, if the mechanisms related to reduced disease severity
are driven by cellular immune responses, there could be additional clinical
benefits to the HD vaccine that cannot be learned from the data we used.

Finally, our results do not preclude a stronger beneficial effect of a higher
vaccine dose. Assuming that a higher dose (higher than the current dose in the
HD vaccine) would further enhance immunogenicity, we could see a stronger effect
which is consistently elevated from the SD vaccine. Since the current HD vaccine
seems to be tolerated well, further studies on optimal dosing in elderly
individuals might open avenues for improved vaccine design using the relatively
cheap split-inactivated vaccine vector.

# Acknowledgments

We would like to acknowledge Michael Carlock (Cleveland Clinic, Port St. Lucie,
FL, USA) for assistance with obtaining data.

**Conflicts of interest**: None of the authors declare any conflicts of interest.

**Funding sources**: YG was supported by the Start-up Grant from the University of Southern Mississippi and NIH contract 75N93019C00052. 
TMR is supported by the Georgia Research Alliance as an Eminent Scholar. 
AH received partial support from NIH grants/contracts U01AI150747, R01AI170116, 75N93019C00052 and 75N93021C00018. 
YS received partial support from NIH grants/contracts R35GM146612, R01AI170116 and 75N93019C00052.
The funders had no role in the study design, data collection and analysis, decision to publish, or preparation of the manuscript.

**Author contributions**:

* WZB: conceptualization, methodology, software, formal analysis, data curation, visualization, writing (original draft)
* YG: conceptualization, methodology, software, formal analysis, data curation, writing (original draft
* JHK: methodology
* HH: writing (review and editing), validation
* SMH: methodology, writing (review and editing), validation
* ALS: writing (review and editing), validation
* WC: writing (review and editing), validation
* YS: methodology, writing (review and editing)
* JB: writing (review and editing)
* PGT: writing (review and editing)
* TMR: data collection, data curation, writing (review and editing)
* AH: conceptualization, data curation, writing (review and editing), supervision, funding acquisition

**Data availability statement**: an archive containing all of the code files and data
necessary to reproduce our results is located on Zenodo at this link:
[https://doi.org/10.5281/zenodo.12666976](https://doi.org/10.5281/zenodo.12666976).

# References

::: {#refs}
:::

<!-- END OF FILE -->
